---
title: "Part 4 Models"
format: html
execute:
  echo: true
  warning: false
  message: false
---


This analysis was conducted using R and the tidymodels framework to evaluate predictive models for lung cancer risk. The workflow includes data preprocessing, model training, evaluation, and comparison across multiple models.

To ensure reproducibility:

- Data were split into training (75%) and testing (25%) sets using stratified sampling based on the outcome variable.
- Models were trained exclusively on the training data and evaluated on an independent test set.
- Model performance was assessed using accuracy, sensitivity, specificity, and area under the ROC curve (AUC).
- 5-fold cross-validation was used to estimate model stability.
- All code, data paths, and outputs are organized using a structured project directory via the `here()` package.

The analysis includes three models:
1. Age + Smoking Status
2. Age + Pack-Years (smoking exposure over time)
3. Age + Pack-Years + COPD

In addition, a LASSO model was implemented to evaluate the effect of regularization and compare predictive performance with the logistic regression models.

All results, tables, and figures can be reproduced by running this document from start to finish.

```{r}
# Load required libraries 
library(tidyverse)
library(tidymodels)
library(here)

set.seed(1234)
```


```{r}
lung_data <- readr::read_csv(
  here("data", "processed-data", "lung_clean.csv")
)
```


```{r}
# Inspect structure and basic distributions
glimpse(lung_data)
table(lung_data$lung_cancer_risk)
table(lung_data$smoker)
table(lung_data$copd)
```

```{r}
# Convert categorical variables to factors for modeling
lung_data <- lung_data %>%
  mutate(
    lung_cancer_risk = factor(lung_cancer_risk),
    smoker = factor(smoker),
    copd = factor(copd)
  )
```

```{r}
glimpse(lung_data)
```


```{r}
set.seed(1234)
# Split data into training (75%) and testing (25%) sets
data_split <- initial_split(
  lung_data,
  prop = 0.75,
  strata = lung_cancer_risk
)

train_data <- training(data_split)
test_data  <- testing(data_split)
```


```{r}
nrow(train_data)
nrow(test_data)

table(train_data$lung_cancer_risk)
table(test_data$lung_cancer_risk)
```


```{r}
# Define logistic regression model for classification
log_model <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")
```


```{r}
# Build workflow using age and smoking status as predictors
log_workflow <- workflow() %>%
  add_model(log_model) %>%
  add_formula(lung_cancer_risk ~ age + smoker)
```


```{r}
# Fit model on training data
log_fit <- fit(log_workflow, data = train_data)
```


```{r}
tidy(log_fit)
```


```{r}
# Generate class predictions and probabilities on test data
test_pred_class <- predict(log_fit, new_data = test_data)
test_pred_prob  <- predict(log_fit, new_data = test_data, type = "prob")
```


```{r}
test_results <- bind_cols(
  test_data %>% select(lung_cancer_risk),
  test_pred_class,
  test_pred_prob
)
```


```{r}
head(test_results)
```


```{r}
# Evaluate model performance using accuracy and confusion matrix
accuracy(test_results, truth = lung_cancer_risk, estimate = .pred_class)
conf_mat(test_results, truth = lung_cancer_risk, estimate = .pred_class)
```


```{r}
# Compute sensitivity and specificity
sens(test_results, truth = lung_cancer_risk, estimate = .pred_class)
spec(test_results, truth = lung_cancer_risk, estimate = .pred_class)
```


```{r}
# Compute AUC
roc_auc(
  test_results,
  truth = lung_cancer_risk,
  .pred_Yes,
  event_level = "second"
)
```

```{r}
# Generate ROC curve data and plot
roc_curve_data <- roc_curve(
  test_results,
  truth = lung_cancer_risk,
  .pred_Yes,
  event_level = "second"
)

autoplot(roc_curve_data)
```

```{r}
# Perform 5-fold cross-validation on training data
cv_folds <- vfold_cv(
  train_data,
  v = 5,
  strata = lung_cancer_risk
)

cv_folds
```


```{r}
# Fit model across folds and compute metrics
cv_results_model1 <- fit_resamples(
  log_workflow,
  resamples = cv_folds,
  metrics = metric_set(accuracy, sens, spec)
)
```


```{r}
collect_metrics(cv_results_model1)
```


```{r}
# Build second model using cumulative smoking exposure (pack-years)
log_workflow_pack <- workflow() %>%
  add_model(log_model) %>%
  add_formula(lung_cancer_risk ~ age + pack_years)
```


```{r}
# Fit model
log_fit_pack <- fit(log_workflow_pack, data = train_data)
```


```{r}
tidy(log_fit_pack)
```


```{r}
# Generate predictions for Model 2
test_pred_class_pack <- predict(log_fit_pack, new_data = test_data)
test_pred_prob_pack  <- predict(log_fit_pack, new_data = test_data, type = "prob")
```


```{r}
# Combine results
test_results_pack <- bind_cols(
  test_data %>% select(lung_cancer_risk),
  test_pred_class_pack,
  test_pred_prob_pack
)
```


```{r}
# Evaluate performance
accuracy(test_results_pack, truth = lung_cancer_risk, estimate = .pred_class)

sens(test_results_pack, truth = lung_cancer_risk, estimate = .pred_class)

spec(test_results_pack, truth = lung_cancer_risk, estimate = .pred_class)
```


```{r}
roc_auc(
  test_results_pack,
  truth = lung_cancer_risk,
  .pred_Yes,
  event_level = "second"
)
```


```{r}
roc_curve_data_pack <- roc_curve(
  test_results_pack,
  truth = lung_cancer_risk,
  .pred_Yes,
  event_level = "second"
)

autoplot(roc_curve_data_pack)
```


```{r}
cv_results_model2 <- fit_resamples(
  log_workflow_pack,
  resamples = cv_folds,
  metrics = metric_set(accuracy, sens, spec)
)

collect_metrics(cv_results_model2)
```

```{r}
acc_m1  <- accuracy(test_results, truth = lung_cancer_risk, estimate = .pred_class) %>% pull(.estimate)
sens_m1 <- sens(test_results, truth = lung_cancer_risk, estimate = .pred_class) %>% pull(.estimate)
spec_m1 <- spec(test_results, truth = lung_cancer_risk, estimate = .pred_class) %>% pull(.estimate)
auc_m1  <- roc_auc(test_results, truth = lung_cancer_risk, .pred_Yes, event_level = "second") %>% pull(.estimate)

acc_m2  <- accuracy(test_results_pack, truth = lung_cancer_risk, estimate = .pred_class) %>% pull(.estimate)
sens_m2 <- sens(test_results_pack, truth = lung_cancer_risk, estimate = .pred_class) %>% pull(.estimate)
spec_m2 <- spec(test_results_pack, truth = lung_cancer_risk, estimate = .pred_class) %>% pull(.estimate)
auc_m2  <- roc_auc(test_results_pack, truth = lung_cancer_risk, .pred_Yes, event_level = "second") %>% pull(.estimate)
```

```{r}
cv_m1 <- collect_metrics(cv_results_model1)
cv_m2 <- collect_metrics(cv_results_model2)
```

```{r}
cv_acc_m1  <- cv_m1 %>% filter(.metric == "accuracy") %>% pull(mean)
cv_sens_m1 <- cv_m1 %>% filter(.metric == "sens") %>% pull(mean)
cv_spec_m1 <- cv_m1 %>% filter(.metric == "spec") %>% pull(mean)

cv_acc_m2  <- cv_m2 %>% filter(.metric == "accuracy") %>% pull(mean)
cv_sens_m2 <- cv_m2 %>% filter(.metric == "sens") %>% pull(mean)
cv_spec_m2 <- cv_m2 %>% filter(.metric == "spec") %>% pull(mean)
```

```{r}
# Create summary table comparing both models (test + CV performance)
model_comparison <- tibble(
  Model = c("Age + smoker", "Age + pack_years"),
  Test_Accuracy = c(acc_m1, acc_m2),
  Test_Sensitivity = c(sens_m1, sens_m2),
  Test_Specificity = c(spec_m1, spec_m2),
  Test_AUC = c(auc_m1, auc_m2),
  CV_Accuracy = c(cv_acc_m1, cv_acc_m2),
  CV_Sensitivity = c(cv_sens_m1, cv_sens_m2),
  CV_Specificity = c(cv_spec_m1, cv_spec_m2)
)

model_comparison
```


```{r}
# Save full comparison table
write.csv(
  model_comparison,
  here("results", "tables", "model_comparison.csv"),
  row.names = FALSE
)
```

```{r}
model_comparison_clean <- model_comparison %>%
  mutate(
    Model = c("Age + Smoking Status", "Age + Pack-Years")
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
# Save cleaned (rounded) version for reporting
write.csv(
  model_comparison_clean,
  here("results", "tables", "model_comparison_clean.csv"),
  row.names = FALSE
)
```

```{r}
model_summary_table <- tibble(
  Model = c("Age + Smoking Status", "Age + Pack-Years"),
  Accuracy = c(acc_m1, acc_m2),
  Sensitivity = c(sens_m1, sens_m2),
  Specificity = c(spec_m1, spec_m2),
  AUC = c(auc_m1, auc_m2)
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
# Save final summary table as RDS
saveRDS(
  model_summary_table,
  here("results", "tables", "model_summary_table.rds")
)
```

```{r}
model_summary_table
```


```{r}
# Save ROC curves for each model and combined comparison

roc_plot_m1 <- autoplot(roc_curve_data)

roc_plot_m1

ggsave(
  filename = here("results", "figures", "roc_model1.png"),
  plot = roc_plot_m1,
  width = 6,
  height = 5
)

roc_plot_m2 <- autoplot(roc_curve_data_pack)

roc_plot_m2

ggsave(
  filename = here("results", "figures", "roc_model2.png"),
  plot = roc_plot_m2,
  width = 6,
  height = 5
)

roc_curve_data$model <- "Model 1"
roc_curve_data_pack$model <- "Model 2"

roc_combined <- bind_rows(roc_curve_data, roc_curve_data_pack)

ggplot(roc_combined, aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_line(linewidth = 1) +
  geom_abline(lty = 2) +
  labs(title = "ROC Curve Comparison", color = "Model") +
  theme_minimal()

ggsave(
  here("results", "figures", "roc_comparison.png"),
  width = 6,
  height = 5
)
```


```{r}
# Model 3: include COPD
log_workflow_copd <- workflow() %>%
  add_model(log_model) %>%
  add_formula(lung_cancer_risk ~ age + pack_years + copd)

# Fit model
log_fit_copd <- fit(log_workflow_copd, data = train_data)
```

```{r}
tidy(log_fit_copd)
```


```{r}
test_pred_class_copd <- predict(log_fit_copd, new_data = test_data)
test_pred_prob_copd  <- predict(log_fit_copd, new_data = test_data, type = "prob")

test_results_copd <- bind_cols(
  test_data %>% select(lung_cancer_risk),
  test_pred_class_copd,
  test_pred_prob_copd
)
```


```{r}
# Check results
head(test_results_copd)
```


```{r}
# Evaluate test-set performance

# Accuracy
accuracy(test_results_copd, truth = lung_cancer_risk, estimate = .pred_class)

# Sensitivity
sens(test_results_copd, truth = lung_cancer_risk, estimate = .pred_class)

# Specificity
spec(test_results_copd, truth = lung_cancer_risk, estimate = .pred_class)

# AUC
roc_auc(
  test_results_copd,
  truth = lung_cancer_risk,
  .pred_Yes,
  event_level = "second"
)
```


```{r}
# Confusion matrix
conf_mat(test_results_copd, truth = lung_cancer_risk, estimate = .pred_class)
```

```{r}
cv_results_model3 <- fit_resamples(
  log_workflow_copd,
  resamples = cv_folds,
  metrics = metric_set(accuracy, sens, spec)
)

collect_metrics(cv_results_model3)
```

```{r}
# Extract Model 3 test metrics
acc_m3  <- accuracy(test_results_copd, truth = lung_cancer_risk, estimate = .pred_class) %>% pull(.estimate)
sens_m3 <- sens(test_results_copd, truth = lung_cancer_risk, estimate = .pred_class) %>% pull(.estimate)
spec_m3 <- spec(test_results_copd, truth = lung_cancer_risk, estimate = .pred_class) %>% pull(.estimate)
auc_m3  <- roc_auc(test_results_copd, truth = lung_cancer_risk, .pred_Yes, event_level = "second") %>% pull(.estimate)

# Create 3-model comparison table (with Sensitivity)
model_comparison_3 <- tibble(
  Model = c("Age + Smoking Status", "Age + Pack-Years", "Age + Pack-Years + COPD"),
  Accuracy = c(acc_m1, acc_m2, acc_m3),
  Sensitivity = c(sens_m1, sens_m2, sens_m3),
  Specificity = c(spec_m1, spec_m2, spec_m3),
  AUC = c(auc_m1, auc_m2, auc_m3)
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

# Show table
model_comparison_3

# Save CSV
write.csv(
  model_comparison_3,
  here("results", "tables", "model_comparison_3.csv"),
  row.names = FALSE
)

# Save RDS
saveRDS(
  model_comparison_3,
  here("results", "tables", "model_comparison_3.rds")
)
```

```{r}
# Add model labels
roc_curve_data$model <- "Model 1"
roc_curve_data_pack$model <- "Model 2"

# Create ROC for Model 3
roc_curve_data_copd <- roc_curve(
  test_results_copd,
  truth = lung_cancer_risk,
  .pred_Yes,
  event_level = "second"
)

roc_curve_data_copd$model <- "Model 3"

# Combine all three
roc_combined <- bind_rows(
  roc_curve_data,
  roc_curve_data_pack,
  roc_curve_data_copd
)

# Plot
ggplot(roc_combined, aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_line(linewidth = 1) +
  geom_abline(lty = 2) +
  labs(title = "ROC Curve Comparison", color = "Model") +
  theme_minimal()

# Save
ggsave(
  here("results", "figures", "roc_comparison_3models.png"),
  width = 6,
  height = 5
)
```

```{r}
# ===============================
# LASSO MODEL (based on Model 3)
# ===============================

library(glmnet)

# Define LASSO model
lasso_model <- logistic_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

# Workflow (same predictors as Model 3)
lasso_workflow <- workflow() %>%
  add_model(lasso_model) %>%
  add_formula(lung_cancer_risk ~ age + pack_years + copd)

# ===============================
# Tuning with Cross-Validation
# ===============================

set.seed(1234)

lasso_tuned <- tune_grid(
  lasso_workflow,
  resamples = cv_folds,   # same CV you already created
  grid = 10,
  metrics = metric_set(accuracy, roc_auc)
)

# View results
collect_metrics(lasso_tuned)
```

```{r}
# Select best penalty based on AUC
best_lasso <- select_best(lasso_tuned, metric = "roc_auc")

best_lasso
```

```{r}
# Finalize workflow with best penalty
final_lasso <- finalize_workflow(lasso_workflow, best_lasso)

# Fit on training data
lasso_fit <- fit(final_lasso, data = train_data)
```

```{r}
# Predictions
lasso_pred_class <- predict(lasso_fit, new_data = test_data)
lasso_pred_prob  <- predict(lasso_fit, new_data = test_data, type = "prob")

# Combine results
lasso_results <- bind_cols(
  test_data %>% select(lung_cancer_risk),
  lasso_pred_class,
  lasso_pred_prob
)

# Metrics
accuracy(lasso_results, truth = lung_cancer_risk, estimate = .pred_class)

roc_auc(
  lasso_results,
  truth = lung_cancer_risk,
  .pred_Yes,
  event_level = "second"
)
```

```{r}
lasso_acc <- accuracy(lasso_results, truth = lung_cancer_risk, estimate = .pred_class) %>% pull(.estimate)

lasso_auc <- roc_auc(
  lasso_results,
  truth = lung_cancer_risk,
  .pred_Yes,
  event_level = "second"
) %>% pull(.estimate)

comparison_final <- tibble(
  Model = c("Logistic (Model 3)", "LASSO"),
  Accuracy = c(acc_m3, lasso_acc),
  AUC = c(auc_m3, lasso_auc)
)

comparison_final
```

```{r}
write.csv(
  comparison_final,
  here("results", "tables", "final_model_comparison.csv"),
  row.names = FALSE
)

saveRDS(
  comparison_final,
  here("results", "tables", "final_model_comparison.rds")
)
```
